---
title: "Case Study 2"
author: "Luke Yates"
date: "`r Sys.Date()`"
output:
  html_document:
    theme:
      version: 5
      bootswatch: cosmo
    toc: true
    toc_float: true
    toc_depth: 3
    highlight: tango
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  dpi = 300,
  collapse = TRUE
)

options(
  width = 120,
  pillar.min_chars = 15,
  pillar.min_title_chars = Inf,
  tibble.print_max = 10
)

# Load required packages
library(cpam)  
library(dplyr)
library(readr)
library(tidyr)
library(stringr)

#"cpam: <u>c</u>hange<u>p</u>oint <u>a</u>dditive <u>m</u>odels"

cpo <- readRDS("../../cpam_paper/output/cpo_crisp_2024_09_13.rds")

```

## About

This tutorial demonstrates the R package `cpam` for the analysis of time series
omics data. It reproduces the results for the second case study presented in the 
manuscript [Shape-constrained, changepoint additive models for time series omics data with cpam]((https://doi.org/10.1101/2024.12.22.630003)).

## Data

The data are oublically available 


## Installation

You can install the package from GitHub using:

```{r installation, eval=FALSE}
devtools::install_github("l-a-yates/cpam")
library(cpam)  
library(dplyr)
library(readr)
library(tidyr)
library(stringr)

```

# Getting Started

## Experimental design and transcript-to-gene mapping
First we create the experimental design which must have at least the following columns: 
time, sample, and path. The rep column is optional and is used here to generate the 
sample names. We have already run kallisto to quantify transcript abundances (with 100 bootstraps),
and here the path column contains the path to the abundance file for each sample.
```{r experimental design}
ed <- 
  expand_grid(time = c(0,30,60,67.5,75,90,120), rep = 1:3) %>%
  mutate(sample = paste0("pc_t",time,"_r",rep),
         path = paste0("case_studies/crisp/data/kallisto/",sample,"/abundance.h5"))

ed

```
The transcript-to-gene mapping for Arabidopsis thaliana can be downloaded from
TAIR (...). We load this now

```{r, eval=FALSE}
t2g <- readr::read_csv("case_studies/crisp/data/t2g.csv")
```

## Fitting the models
To fit the models, we first prepare the cpam object, then compute p-values, 
estimate the changepoint, and select the shape of each transcript. The last step 
takes the longest (here just under 13 minutes) but it is worth the wait to be able to
visualise and cluster the transcripts by shape.

```{r fitting the model, eval = F}
  cpo <- prepare_cpam(exp_design = ed,
                      model_type = "case-only",
                      t2g = t2g,
                      import_type = "kallisto",
                      num_cores = 4)
  cpo <- compute_p_values(cpo) # 1:52
  cpo <- estimate_changepoint(cpo) # 6:32 secs
  cpo <- select_shape(cpo) # 12:54 secs
    
```
We can look at a summary of the fitted cpam object 
```{r}
cpo
```
Although it cannot be launched here on the tutorial page, you can access the Shiny app
to visualise the results by running the following code `visualise(cpo)`.

## Tables of Results
The results of the analysis are summarised using the `results` function.
```{r feature-1}
results(cpo)
```
The generated results list can be filtered by specifying minimum counts, minimum
log-fold changes, and maximum p-values. For example, to return only the transcripts
with a log-fold change greater than 2, at least 50 counts, and a p-value less than 0.01, we can run
```{r filtered-results}
results(cpo, min_count = 10, min_lfc = 1, p_threshold = 0.01)
```


## Plotting Genes and Transcripts

A single gene can be plotted using the `plot_cpam` function. Here we plot the gene AT1G64140

```{r plot1}
plot_cpam(cpo, gene_id = "AT1G64140")
```
The substitle shows (0,tp) indicating a changepoint at time point 0 (or no changepoint) and an unconstained 'tp' (thinplate) shape. This selection of 'tp' suggest that the trend for this gene does not conform to one of the simpler shape types that cpam uses. To force the cpam to choose among the simpler forms, we set `shape_type = "shape2"` in the `plot_cpam` function. 
```{r plot1b}
plot_cpam(cpo, gene_id = "AT1G64140",shape_type = "shape2")
```
Here a concave shape ('cv') is chosen, and we can see this deviates from the data more that the 
'tp' shape.

Next we plot a gene with multiple transcripts, some of which have changepoints different
from 0.

```{r}
plot_cpam(cpo, gene_id = "AT1G28610")
```
The first transcript has a changepoint at 30 mins and the second at 0. Both have an 
unconstrained shape. Changepoint, shape and other results from the fitted models can also be
extracted manually from the cpam object. For example, to extract the shape of the transcripts
```{r}
cpo$shapes %>% filter(str_starts(target_id,"AT1G28610"))
```
Lastly, we plot the two remaining genes shown in the manuscript, AT4G34590 and AT3G23280. 

```{r}
plot_cpam(cpo, gene_id = "AT4G34590")
```

```{r}
plot_cpam(cpo, gene_id = "AT3G23280")
```


## Clusters
The results function can be used to generate clusters according to selected filters. Here we
generate clusters of transcripts with at least 100 counts, a log-fold change greater than 1, and a p-value less than 0.01. The `plot_cluster` function can then be used to visualise the clusters which we do here for targets with changepoints at 30 mins and the 'micv' (montonic increasing concave) shape.

```{r clusters, cache=TRUE}
res <- results(cpo, min_count = 100, min_lfc = 1, p_threshold = 0.01)
plot_cluster(cpo, res, changepoints = 30, shapes = "micv")
```
Clustering can be further refined based on, for example, the rate at which the above transcripts
attain their maximum values. We illustrate refinements such as this in another case study [here](url).

# Session Info

```{r session-info}
sessionInfo()
```
