---
title: "Human Embryo Case Study"
author: "Luke Yates"
date: "`r Sys.Date()`"
bibliography: references.bib
output:
  html_document:
    theme:
      version: 5
      bootswatch: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    highlight: tango
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  dpi = 300,
  collapse = TRUE
)

options(
  width = 150,
  pillar.min_chars = 15,
  pillar.min_title_chars = Inf,
  tibble.print_max = 10
)

# Load required packages
library(cpam)  
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(purrr)
#"cpam: <u>c</u>hange<u>p</u>oint <u>a</u>dditive <u>m</u>odels"

cpo <- readRDS("../../cpam_paper/output/cpo_tx_torre_2024_12_11.rds")
directory_path <- "../../cpam_paper/case_studies/torre/data/"
```

## About

This tutorial demonstrates the R package `cpam` for the analysis of time series
omics data. It serves an introduction to the package and reproduces the results 
for the first case study presented in the accompanying 
[manuscript](https://doi.org/10.1101/2024.12.22.630003) by @Yates2024. 
The second case study using Arabidopsis thaliana time series data is available [here](https://raw.githack.com/l-a-yates/cpam_manuscript/main/R/crisp.html).


## Data

This tutorial uses time-series RNA-seq data tracking human embryo development across multiple stages. The data are publicly available for download [here](https://denis-torre.github.io/embryo-transcriptome/). 
The data consist of transcript counts for the human preimplantation embryo
and are described in original data [manuscript](https://doi.org/10.1038/s41467-023-42558-y) by @Torre2023.
The experiment captures transcriptional changes across six early developmental stages: zygote, 2-cell, 4-cell, 8-cell, morula, and blastocyst. These time points span a major developmental transition - embryonic genome activation (EGA) that occurs after the 4-cell stage – and captures major transcriptional transitions in development.  

## Installation

You can install `cpam` from GitHub using:

```{r installation, eval=FALSE}
devtools::install_github("l-a-yates/cpam")
```

# Getting started

## Load packages 

```{r loading, eval=FALSE}
library(cpam)  
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(purrr)

directory_path <- "torre/data/"
```

## Load data and transcript-to-gene mapping

We load the transcript counts and the transcript-to-gene mapping. The transcript counts
are stored in a matrix where the rows are the transcript IDs and the columns are the samples.

```{r load-data}
sample_names <- readr::read_tsv(
  paste0(directory_path, "human_embryo-transcript_counts.tsv"),
  n_max = 1, 
  col_names = F
) %>%
  t %>% as.vector()

counts <- readr::read_tsv(
  paste0(directory_path, "human_embryo-transcript_counts.tsv"),
  skip = 1,
  col_names = c("target_id", sample_names)
)

count_matrix <- counts %>% dplyr::select(-target_id) %>% as.matrix
rownames(count_matrix) <- counts %>% pull(target_id)

# convert to tibble (for display only)
as.data.frame(count_matrix) %>% head %>% tibble::rownames_to_column("target_id")

```

The transcript-to-gene mapping is a data frame with two columns, target_id and gene_id.
```{r t2g}
transcriptome <- readr::read_tsv(paste0(directory_path,"Human_embryo_isoform_transcriptome_GRCh38-isoform_classification.tsv"))

t2g <- transcriptome %>%
  select(target_id = transcript_id, gene_id) %>%
  filter(target_id %in% rownames(count_matrix))

t2g
```


## Experimental design
The experimental design is a tibble containing at least the following columns: 
time and sample. The stage column is optional and used here to match sample
names with the time points.
```{r experimental-design}

stage_to_time <- tibble(stage = c("2PN","2C","4C","8C","morula","blastocyst"),
                        time = c(17/24,1.5,2,3,4.5,5.5) %>% round(2))

ed <-
  tibble(sample = colnames(count_matrix)) %>%
  mutate(stage = str_match(sample,"_(.*?)_")[,2]) %>%
  left_join(stage_to_time, by = "stage")

ed
```

## Fitting the models
To fit the models, we first prepare the `cpam` object, then compute p-values, 
estimate the changepoint, and select the shape for each transcript. 
For this analysis, the count data are supplied as a matrix (see this [case study](https://raw.githack.com/l-a-yates/cpam_manuscript/main/R/crisp.html) for importing data directly from common quantification software). We filter out transcripts
that do not have more than 20 reads in at least 3/5 of the samples. 

```{r fitting-the-model, eval = F}
cpo <- prepare_cpam(
  exp_design = ed,
  count_matrix = count_matrix,
  model_type = "case-only",
  filter_fun_args = list(min_reads = 20, min_prop = 3/5),
  t2g = t2g,
  num_cores = 4
)
cpo <- compute_p_values(cpo) # 00:03:37 hms
cpo <- estimate_changepoint(cpo) # 00:23:26 hms
cpo <- select_shape(cpo) #  00:37:28 hms
    
```
For larger transcriptomes, such as this human transciptome with many additional transcripts beyond the standard reference, the model fitting can take a long time. Here, using 4 cores, we were able to estimate $p$-values
in 3 minutes, changepoints in 23 minutes, and shape selections in 37 minutes. The latter
stages, although slower to compute, are worth the wait for the inferences they provide.


We can look at a summary of the fitted cpam object 
```{r print-cpo, results='hold'}
cpo
```
If you run the code on your own computer, you can launch the Shiny app
on your own computer to visualise the results interactively using `visualise(cpo)`.


## Results
The results of the analysis are summarised using the `results` function. Here
we use the default settings for log-fold change and p-value thresholds. 

```{r results}


res <- results(
  cpo,
  min_lfc = 0,
  p_threshold = 0.05,
  add_counts = T,
  add_lfc = F
)

res
```
Using the results table, we can plot the number of differentially expressed genes
at each time point. 
```{r degs-by-timepoint}
res %>%
  pull(cp) %>%
  table %>%
  {tibble(cp = names(.), n = as.numeric(.))} %>%
  ggplot(aes(cp, n)) +
  geom_point() +
  geom_line(aes(group = 1)) +
  theme_classic()
```
There is an obvious peak in differentially expressed genes at the 4C stage (t = 2.0 hours), corresponding to the
embryonic genome activation (EGA).

## Plotting

To plot the fitted models for a specific gene, we can use the `plot_cpam` function.
```{r plot-gene-1a}
g <- "ENSG00000141378"
plot_cpam(
  cpo,
  gene_id = g,
  facet = T, # plot each transcript on a separate facet
  remove_null = T, # do show transcripts with null trajectories
  common_y_scale = F # do not use a common y scale
) 
```
The `facet` argument can be set to FALSE to plot all the fits on the same plot.

```{r plot-gene-1b}
g <- "ENSG00000141378"
plot_cpam(cpo,
          gene_id = g,
          facet = F, # plot each transcript in a common plot
          remove_null = T)
```

## Customising the plot
If you require more customisation than the `plot_cpam` function allows, you can extract the
model fits and plot their predictions manually. Here we make a logged version 
of the above plot with some extra customisations such as changepoint ticks.

```{r plot-gene-manual}

# set gene
g <- "ENSG00000141378"

# set logged
logged <- T
fun <- identity
if(logged) fun <- function(x) log(x + 0.5)
scls <- if_else(logged, "free_x", "free")

# get data
obs <-
  cpo$data_long %>% 
  filter(gene_id == g) %>%
  transmute(target_id, time, counts = fun(counts_raw/norm_factor + 0.5))
cps <- cpo$changepoints %>% filter(target_id %in% unique(obs$target_id))

# set colours
tx_cols <- c(RColorBrewer::brewer.pal(7, name = "Dark2"))[-4]

# plot
plot_cpam(cpo, gene_id = g, return_fits_only = T) %>%
  map_dfr( ~ cpam:::predict_cpgam(.x, logged = logged), .id = "target_id") %>%
  ggplot(aes(time, counts)) +
  scale_x_continuous(breaks = round(sort(cpo$times), 1),
                     labels = c("0.7", "", "2.0", "3.0", "4.5", "5.5")) +
                     #labels = c("1C","1C","4C","8C","M","B")) +
                     geom_line(aes(col = target_id)) +
                       geom_ribbon(aes(ymin = q_lo, ymax = q_hi, fill = target_id), alpha = 0.15) +
                       facet_wrap( ~ target_id, scales = scls) +
                       geom_point(data = obs, aes(time, counts, col = target_id)) +
                       scale_fill_manual(values = tx_cols, aesthetics = c("fill", "color")) + 
                       theme_classic() +
                       theme(
                         legend.position = "bottom",
                         plot.subtitle = element_text(
                           hjust = 0,
                           face = "plain",
                           margin = margin(b = 10)
                         ),
                         strip.background = element_blank(),
                         strip.text = element_blank(),
                         axis.line.x = element_line(color = "black"),
                         axis.ticks.x = element_line(),
                         panel.spacing.y = unit(1, "lines"),
                         panel.spacing.x = unit(1, "lines"),
                         axis.text.x = element_text(angle = 0, hjust = 0.5)
                       ) +
                       labs(
                         y = "logged counts",
                         x = "time (days)",
                         col = NULL,
                         fill = NULL,
                         subtitle = paste0(g)
                       ) +
                       guides(color = guide_legend(byrow = TRUE), fill = guide_legend(byrow = TRUE)) +
                       geom_point( # add ticks for changepoints
                         data = cps,
                         aes(x = cp_1se, y = -Inf),
                         shape = 17,
                         color = "grey20",
                         size = 2.5
                       )
                     
                     
```

## Clusters

We can identify clusters of targets that have similar expression profiles. In the 
first instance, we can use the results table with the function `plot_cluster`

```{r plot-cluster-1, cache = TRUE}
# filter to keep number of targets reasonable
res <- results(cpo, min_lfc = 2, min_count = 1000)

plot_cluster(cpo, res, changepoints = 2.0, shapes = "micv")

```

There are more than 4000 targets in this subset, and such a large number can obscure subsets of
targets with different expression profiles, for example with differing rates of increase.

When the default clustering categories yield a large numbers of targets, it is
useful to be able define custom clusters. Here we define clusters based on the rate of the change in expression between the 3C and 4.5C stages.

To contain the scale of the plot, we filter the results table to remove a handful of
large log-fold changes above 10. We define the rate to be ratio of the log-fold changes between the 3C and 4.5C stages and choose three thresholds to define the clusters.
```{r res-cluster}

res_cluster <-
  res %>% 
  filter(lfc.5.5<10) %>% # remove extreme values
  filter(cp %in% 2, shape %in% "micv") %>% # select the 4C stage and the micv shape
  mutate(rate = lfc.3/lfc.4.5) %>% # calculate the speed
  mutate(cluster = cut(rate,breaks = c(0, 0.6, 0.75, 0.87,10), labels = paste0("cluster ",1:4)))

res_cluster
```
This yields 1282 targets across four clusters. Next we generate the plot data which takes a minute to compute.

```{r cluster-data, cache = TRUE}
cluster_plot_data <-
  res_cluster %>%
  pull(target_id) %>%
  set_names() %>%
  map_dfr(~plot_cpam(cpo, target_id = .x, return_fits_only = T) %>%
            cpam:::predict_lfc(),.id = "target_id")

```
Finally, we plot the clusters.
```{r cluster-plot, cache = TRUE}

cluster_plot_data %>%
  left_join(res_cluster %>% select(target_id,speed), by = "target_id") %>%
  ggplot(aes(x = time,y = lfc, group = target_id, col = speed)) +
  geom_line(alpha = 0.2) +
  scale_x_continuous(breaks = round(sort(cpo$times),1),
                     labels = c("0.7","","2.0","3.0","4.5","5.5")) +
  scale_color_manual(values = blues9[5:8]) +
  facet_wrap(~ speed, nrow = 1) +
  theme_classic() +
  theme(legend.position = "none",
        strip.background = element_blank(),
        panel.spacing.x = unit(1.5, "lines"),
        strip.text = element_text(size = 10, family = "sans")) +
  labs(title = NULL, subtitle = NULL,
       x = "time (days)", y = "log-fold change")

```

For further, such as GO term enrichment, we can easily extract tables of clustered
targets. To look at the numbers of targets in each cluster.

```{r}
res_cluster %>% pull(cluster) %>% table
```
To match targets with their rate and cluster.
```{r}
res_cluster %>% select(target_id,rate,cluster)
```

To look at the targets in cluster 1.
```{r }
res_cluster %>% select(target_id,rate) %>% 
filter(rate == "cluster 1") %>% pull(target_id) %>% head

```

## References
